Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-05-15 publication_materials/ms_rbc.tex
Evan's feedback

\documentclass[reprint,amsmath,amssymb,aps]{revtex4-1}

@@

\newcommand\eps{\varepsilon}
\renewcommand\L {\mathcal{L}}

@@

\section{Introduction}
Rayleigh-B\'enard convection plays a foundational role in astrophysical and geophysical settings.
The resulting buoyancy-driven flows regulate heat transfer and generate large-scale vortices \cite{Couston}.
Turbulent convection, which is associated with large Rayleigh numbers $\Ra$, is difficult to simulate. 
State of the art simulations performed by \cite{Zhu_2018} have reached $\Ra \sim 10^{14}$ but estimates for the sun's convective zone and earth's interior are $\Ra \sim 10^{16}-10^{20}$ and $\Ra \sim 10^{20}-10^{30}$ respectively \cite{Ossendrijver,Gubbins_2001}. 
The scaling behavior of the Nusselt number $\Nu \sim \Ra^{\beta}$ in the asymptotic ultimate regime is of particular interest.
There exists a substantial body of work pertaining to this specific topic with no general consensus \cite{Malkus_1954, Howard_1966, Kraichnan, Spiegel, Castaing, Grossman, Ahlers}.

Absent solid evidence from direct numerical simulation, other methods have been developed to try to infer large-$\Ra$ behavior or otherwise gain insight.
In the presence of other physical effects (e.g., rotation, magnetic fields), one can sometimes derive an asymptotically consistent set of reduced equations \cite{Julien2007, Julien2012}.
Reduced models are potentially useful in this context because they may allow us to study the problem with less expensive computations.
Another approach relates to unstable exact coherent states (ECS) which are steady solutions to the full nonlinear problem \cite{Waleffe, Sondak, Wen, chini_cells}. 
Simulations and analysis performed by \cite{Yalniz, Cvitanovic} suggest that chaotic solution trajectories might ``visit'' these ECS.
Should that be the case, it is crucial that we discover and classify such equilibria.

Others have turned to studying quasilinear systems.
The quasilinear approximation starts with a decomposition of all variables into a background and perturbations about this background.
This approximation neglects the nonlinear interactions between perturbations which modify the perturbations themselves \cite{marston2016}.
This renders the perturbation equations linear.
Although the quasilinear approximation greatly simplifies the problem, an additional condition must be imposed to determine the amplitude of the perturbations.
In \cite{Beaume_2015}, researchers compute ECS in parallel shear flows by deriving and solving a quasilinear formulation of the Navier-Stokes equations via multi-scale asymptotic arguments. 
They assume the background velocity evolves on a slow time-scale, and, to determine the perturbation amplitudes, they require marginal stability at each timestep.
A similar strategy is employed by \cite{michel_chini_2019} to study acoustic streaming.
In that work, an analytic expression for the first-order perturbation's amplitude is found by deriving a solvability condition.

@@

\section{Model Setup}\label{sec:model}
We begin with the Boussinesq approximation for Rayleigh-Bénard Convection, nondimensionalized on the freefall time-scale. 
The domain $\mathcal{D}$ is 2-dimensional, rectangular, and horizontally periodic with spatial dimensions $0 \leq x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two flat boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1/2$ and $-1/2$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\vec{u} = u \hat{x} + w \hat{z} = \vec{0}$ at $z = \pm 1/2$, where $\hat{x}, \hat{z}$ are the unit vectors in the $x$ and $z$ directions. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \vec{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \vec{u}}{\partial t} + \vec{u} \cdot \nabla \vec{u} &= - \nabla p + T \hat{z} + \mathcal{R} \nabla^2 \vec{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \vec{u} \cdot \nabla T &= \mathcal{P} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
%For completeness, we specify a final boundary condition $p = p_0$ at $z = - 1/2$. 
Any system of this form can be characterized by its dimensionless Rayleigh number $\Ra = \frac{g\alpha L^3 \Delta T}{\nu \kappa}$ and Prandtl number $\Pr = \frac{\nu}{\kappa}$, where $g, \, \alpha, \, L_z, \, \Delta T, \nu, \kappa$ are the gravitational acceleration, coefficient of thermal expansion, domain height, opposed temperature difference, kinematic viscosity, and thermal diffusivity respectively. 
In this paper, we fix $\Pr = 1$.
For convenience, we define
\begin{equation}
\mathcal{R} = \sqrt{\frac{\Pr}{\Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{\Pr \Ra}}.
\end{equation}

To derive the quasilinear form, we posit that an arbitrary field $f$ can be represented as the sum of a mean profile (denoted by $\bar{f}$) and a perturbation function (denoted by $f'$).
\begin{align}
    \vec{u}(x, z, t) &= \vec{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \bar{T}(z, t) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \bar{p}(z, t) +  p'(x, z, t) \label{EQ:reynolds_dc_p}.
\end{align}
where the mean-velocity components vanish due to incompressibility and symmetry. Perturbations are defined to have no horizontal-average
\begin{equation}
    \langle f'(x, z, t) \rangle_x \equiv \int_{0}^4 f'(x, z, t) dx = 0.
\end{equation}
%Assuming the existance of nontrivial solutions, we also fix the volume-average over the entire domain
%\begin{equation}
%  \langle T'^2 \rangle_{\mathcal{D}} = 1
%\end{equation}
%thereby alleviating any ambiguity in $A(t)$. 
Substituting \eq{EQ:reynolds_dc_T} into \eq{EQ:motion3} and taking the horizontal-average reduces the system to a simple initial value problem (IVP) for $\bar{T}$
\begin{equation}
  \frac{\partial \bar{T}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = \mathcal{P} \frac{\partial^2 \bar{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
with associated boundary conditions $\bar{T}(-1/2, t) = 1/2$ and $\bar{T}(1/2, t) = -1/2$. It should be noted that we could obtain a similar IVP for $u$ by breaking symmetry and considering some nontrivial mean horizontal flow $\bar{u}(z, t)$. However we must have $\bar{w}(z, t) = 0$ due to incompressibility.

Substituting \eq{EQ:reynolds_dc_p} into \eq{EQ:motion2} and taking the horizontal average reveals that the mean pressure field $\bar{p}(z)$ must satisfy
\begin{equation}
    0 = -\frac{\partial \bar{p}}{\partial z} + \bar{T}\label{EQ:p_bar}
\end{equation}

To solve \eq{EQ:T0_IVP} numerically, we need an expression for the perturbations so we can calculate the advective heat flux.
Here we will make the quasilinear approximation, dropping the $\vec{u}'\vec{\cdot}\vec{\nabla}\vec{u}'$ and $\vec{u}'\vec{\cdot}\vec{\nabla}T'$ terms from the evolution equations for the perturbations.
Substituting \eqss{EQ:reynolds_dc_u}{EQ:reynolds_dc_p} into \eqss{EQ:motion1}{EQ:motion3} followed by subtracting \eqs{EQ:T0_IVP}{EQ:p_bar} from the resulting temperature and $\hat{z}$ momentum equations gives
\begin{align}
    \nabla \cdot \vec{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\vec{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \vec{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad u'|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
This is now a linear problem in $\vec{u}'$ and $T'$ which can be solved as an eigenvalue problem.

@@

For each $k_x$, we can assess the stability of the perturbations by solving for the eigenvalue $s$, whose imaginary component $\sigma$ plays the role of an exponential growth rate. 
Positive eigenvalues indicate that the system is unstable to small disturbances of wavenumber $k_x$, while negative eigenvalues indicate stability. 
A complete linear stability analysis requires solution over the full spectrum of wavenumbers. 
The prototypical case is used to demonstrate that the critical Rayleigh number $\Ra_c = 1708$ when $\frac{\partial \bar{T}}{\partial z} = -1$.

To calculate the advective heat flux in equation~\ref{EQ:T0_IVP}, we can sum $\langle w'T' \rangle_x$ from each horizontal wavenumber individually.
In this way, the heat flux from the perturbations influence the evolution of $\bar{T}$.
But the evolution of $\bar{T}$ also influences the perturbations, as equation~\ref{EQ:linear3} depends on $\partial_z \bar{T}$.
Thus, the mean temperature and perturbation fields are coupled.

\section{Perturbation Evolution}\label{sec:evolution}
The linearized system \eqss{EQ:linear1}{EQ:linear3} does not constrain the amplitude of the eigenmodes, $A$.
However, the advective heat flux is proportional to $A^2$, so we need to specify the amplitude in order to solve equation~\ref{EQ:T0_IVP}.
To evolve $\bar{T}$, we assume the perturbations evolve on a much faster time-scale than the mean temperature, as in \cite{michel_chini_2019}.
Stable modes $(\sigma < 0)$ decay away rapidly. 
Unstable modes $(\sigma > 0)$ will not persist on the slow time-scale because the advective term $\langle w'T' \rangle_x$ tends to stabilize $\bar{T}$, thereby creating a negative feedback loop.
Only marginally stable modes can be maintained on the slow time-scale.
Therefore the amplitude $A$ must satisfy
\begin{equation}
    \max_{k_x} \{ \sigma \} = 0.
\end{equation}

For various $\Ra$ and fixed $\Pr = 1$, we seek marginally-stable thermal equilibria (MSTE) satisfying $\frac{\partial \bar{T}}{\partial t} = 0$ according to \eq{EQ:T0_IVP}. 
We employ the \texttt{Dedalus} pseudo-spectral python framework \cite{Dedalus_2020} to solve the eigenvalue problem outlined in Section \ref{sec:model} as well as the time evolution equations \eq{EQ:T0_IVP}.
We represent each field with Chebyshev polynomials and use the 3/2 dealiasing rule to calculate the advective heat flux.
The necessary number of basis functions increases with $\Ra$ as the eigenfunctions include increasingly small-scale features (see Appendix \ref{sec:tables}). 
We use the \texttt{Eigentools} package \cite{Eigentools} to manipulate the eigenfunctions and calculate the advective heat flux $\langle w' T' \rangle_x$.

To initialize our calculations, we construct a marginally-stable initial temperature profile $\bar{T}(z, t=0)$ whose equation is given in appendix~\ref{sec:initial_profile}. 
We use \eq{EQ:T0_IVP} to evolve $\bar{T}(z, t_0)$ into a new marginally-stable profile $\bar{T}(z, t_0 + \Delta t)$.
We use a second-order, two-stage IMEX Runge-Kutta method.
The eigenfunctions and amplitudes are assumed to be constant over the timestep.
We symmetrize $\bar{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each timestep.
This ensures $\bar{T}(z) = -\bar{T}(-z)$.
On every timestep, we calculate the marginally stable eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $\Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\bar{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing $\sigma$. 
    Advection tends to stabilize the system, decreasing $\sigma$. 
    We find the eigenfunction amplitude $A^2$ such that the combination of diffusion and advection yields a new, ``evolved,'' marginally-stable mean temperature profile $\bar{T}(z, t_0 + \Delta t)$. In this case $A^2 \approx 1$ due to approximately equal magnitudes of the diffused and advected eigenvalues.}
    \label{fig:iteration_spectra} 
\end{figure}

Consider a marginally-stable temperature profile $\bar{T}(z, t_0)$.
By definition, its maximum growth rate is 0.
Diffusing $\bar{T}(z, t_0)$ tends to increase its growth rates while ignoring the diffusive term and evolving according to advection tends to decrease its growth rates.
$A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new IVPs
\begin{align}\label{eqn:diffusion_only}
    &\frac{\partial \bar{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \bar{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \bar{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = 0 
\end{align}
where $\bar{T}_{\rm{diff}}$ and $\bar{T}_{\rm{adv}}$ denote to the diffused and advected temperature profiles.
We initialize $\bar{T}_{\rm{diff}}$ and $\bar{T}_{\rm{adv}}$ with $\bar{T}(z, t_0)$, and evolve the equations to $t=t_0+\Delta t$.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then, using the diffused and advected profiles, we compute the eigenvalues $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$ and $t=t_0+\Delta t$.
A preliminary approximation $\Lambda^2_0$ of the appropriate amplitude $A^2$ is given by
\begin{equation}
    A^2 \approx \Lambda^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2$ and $\Lambda^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.

Using the approximate amplitude $\Lambda_0^2$, the background temperature $\bar{T}(z, t_0)$ is then evolved to $t=t_0+\Delta t$ according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
This background temperature is typically close to, but not exactly, marginally stable.
We use Newton's method to find an amplitude $A^2$ such that the maximum growth rate is zero to within $10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e. $\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In section \ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.

@@

Our estimate $\vec{A^2_0}$ does not give a background temperature for which all $N$ modes are exactly marginally stable.
We refine our vector of amplitudes via Newton's method.
This requires calculating the Jacobian matrix 
\begin{equation}
    J = \begin{bmatrix}
        \nabla \sigma_1 (A_1, A_2, ..., A_N) \\
        \nabla \sigma_2 (A_1, A_2, ..., A_N) \\
        \vdots \\
        \nabla \sigma_N (A_1, A_2, ..., A_N) 
    \end{bmatrix}.
\end{equation}
We approximate $J$ via first-order finite differences, which requires an additional $N^2$ sparse eigenvalue solves.
%A once-refined estimate $\vec{A^2_1}$ is then given by
%% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
%\begin{equation}
%    \vec{\tilde{A}^2} \approx \vec{A^2_1} = -J^{-1} \vec{\sigma}(\vec{A^2_0}).
%\end{equation}
%We then use each subsequent guess to adjust the Jacobian via Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We iterate Newton's method until all marginally stable modes have eigenvalues within $10^{-9}$ of zero.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the coupling of $\langle w'T' \rangle_x$ with $\bar{T}$.
Over the course of a large time step, $\bar{T}$ evolves according to \eq{EQ:T0_IVP} and eventually the original eigenfunctions cease to provide a stabilizing influence.
This limits the timestep size we can take.

>>>>

\documentclass[reprint,amsmath,amssymb,aps,nofootinbib]{revtex4-1}

@@

\newcommand\eps{\varepsilon}
\renewcommand\L {\mathcal{L}}
\renewcommand{\citet}[1]{ref.~\cite{#1}}
\renewcommand{\Citet}[1]{Ref.~\cite{#1}}

@@

\section{Introduction}
Rayleigh-B\'enard convection plays a foundational role in astrophysical and geophysical settings.
The resulting buoyancy-driven flows regulate heat transfer and generate large-scale vortices \cite{Couston}.
Turbulent convection, which is associated with large Rayleigh numbers $\Ra$, is difficult to simulate. 
State of the art simulations performed by \cite{Zhu_2018} have reached $\Ra \sim 10^{14}$ but estimates for the sun's convective zone and earth's interior are $\Ra \sim 10^{16}-10^{20}$ and $\Ra \sim 10^{20}-10^{30}$ respectively \cite{Ossendrijver,Gubbins_2001}. 
The scaling behavior of the Nusselt number $\Nu \sim \Ra^{\beta}$ in the asymptotic ultimate regime is of particular interest.
There is a substantial body of work pertaining to this specific topic with no general agreement between numerical simulations and asymptotic theories \cite{Malkus_1954, Howard_1966, Kraichnan, Spiegel, Castaing, Grossman, Ahlers}.

Absent solid evidence from direct numerical simulation, other methods have been developed to try to infer large-$\Ra$ behavior or otherwise gain insight.
In the presence of other physical effects (e.g., rotation, magnetic fields), one can sometimes derive an asymptotically consistent set of reduced equations \cite{Julien2007, Julien2012}.
Reduced models are useful because they allow us to study the problem with less expensive computations.
Another approach relates to unstable exact coherent states (ECS) which are steady solutions to the full nonlinear problem \cite{Waleffe, Sondak, Wen, chini_cells}. 
Simulations and analyses performed by \cite{Yalniz, Cvitanovic} suggest that chaotic solution trajectories might ``visit'' these ECS.
Should that be the case, it is crucial that we discover and classify such equilibria.

Others have turned to studying quasilinear systems.
The quasilinear approximation starts with a decomposition of all variables into a background and perturbations about this background.
This approximation neglects the nonlinear interactions between perturbations which modify the perturbations themselves \cite{marston2016}.
This renders the perturbation equations linear.
Although the quasilinear approximation greatly simplifies the problem, an additional condition must be imposed to determine the amplitude of the perturbations.
In \citet{Beaume_2015}, researchers compute ECS in parallel shear flows by deriving and solving a quasilinear formulation of the Navier-Stokes equations via multi-scale asymptotic arguments. 
They assume the background velocity evolves on a slow time-scale, and, to determine the perturbation amplitudes, they require marginal stability at each timestep.
A similar strategy is employed by \citet{michel_chini_2019} to study acoustic streaming.
In that work, an analytic expression for the first-order perturbation's amplitude is found by deriving a solvability condition.

@@

\section{Model Setup}\label{sec:model}
We begin with the Boussinesq approximation for Rayleigh-Bénard Convection, nondimensionalized on the freefall time-scale. 
The domain $\mathcal{D}$ is 2-dimensional, rectangular, and horizontally periodic with cartesian spatial coordinates $0 \leq x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two flat boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1/2$ and $-1/2$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\vec{u} = u \hat{x} + w \hat{z} = \vec{0}$ at $z = \pm 1/2$, where $\hat{x}, \hat{z}$ are the unit vectors in the $x$ and $z$ directions. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \vec{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \vec{u}}{\partial t} + \vec{u} \cdot \nabla \vec{u} &= - \nabla p + T \hat{z} + \mathcal{R} \nabla^2 \vec{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \vec{u} \cdot \nabla T &= \mathcal{P} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
%For completeness, we specify a final boundary condition $p = p_0$ at $z = - 1/2$. 
Any system of this form can be characterized by its dimensionless Rayleigh number $\Ra = \frac{g\alpha L^3 \Delta T}{\nu \kappa}$ and Prandtl number $\Pr = \frac{\nu}{\kappa}$, where $g, \, \alpha, \, L_z, \, \Delta T, \nu, \kappa$ are the gravitational acceleration, coefficient of thermal expansion, domain height, opposed temperature difference, kinematic viscosity, and thermal diffusivity respectively. 
In this paper, we fix $\Pr = 1$.
For convenience, we define
\begin{equation}
\mathcal{R} = \sqrt{\frac{\Pr}{\Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{\Pr \Ra}}.
\end{equation}

To derive the quasilinear form, we posit that an arbitrary field $f$ can be represented as the sum of a mean profile (denoted by $\overline{f}$) and a perturbation function (denoted by $f'$).
\begin{align}
    \vec{u}(x, z, t) &= \vec{u'}(x, z, t) \label{EQ:reynolds_dc_u}\\
    &= u'(x, z, t)\hat{x} + w'(x, z, t)\hat{z} \\
    T(x, z, t) &= \overline{T}(z, t) + T'(x, z, t) \label{EQ:reynolds_dc_T}\\
    p(x, z, t) &= \overline{p}(z, t) +  p'(x, z, t) \label{EQ:reynolds_dc_p}.
\end{align}
where the mean-velocity components vanish due to incompressibility and symmetry. Perturbations are defined to have no horizontal-average
\begin{equation}
    \langle f'(x, z, t) \rangle_x \equiv \int_{0}^4 f'(x, z, t) dx = 0.
\end{equation}
%Assuming the existance of nontrivial solutions, we also fix the volume-average over the entire domain
%\begin{equation}
%  \langle T'^2 \rangle_{\mathcal{D}} = 1
%\end{equation}
%thereby alleviating any ambiguity in $A(t)$. 
Substituting \eq{EQ:reynolds_dc_T} into \eq{EQ:motion3} and taking the horizontal-average reduces the system to a simple initial value problem (IVP) for $\overline{T}$
\begin{equation}
  \frac{\partial \overline{T}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = \mathcal{P} \frac{\partial^2 \overline{T}}{\partial z^2}, \label{EQ:T0_IVP}
\end{equation}
with associated boundary conditions $\overline{T}(-1/2, t) = 1/2$ and $\overline{T}(1/2, t) = -1/2$. It should be noted that we could obtain a similar IVP for $u$ by breaking symmetry and considering some nontrivial mean horizontal flow $\overline{u}(z, t)$. However we must have $\overline{w}(z, t) = 0$ due to incompressibility.

Substituting \eq{EQ:reynolds_dc_p} into \eq{EQ:motion2} and taking the horizontal average reveals that the mean pressure field $\overline{p}(z)$ must satisfy
\begin{equation}
    0 = -\frac{\partial \overline{p}}{\partial z} + \overline{T}\label{EQ:p_bar}
\end{equation}

To solve \eq{EQ:T0_IVP} numerically, we need an expression for the perturbations so we can calculate the advective heat flux.
Here we will make the quasilinear approximation, dropping the $\vec{u}'\vec{\cdot}\vec{\nabla}\vec{u}'$ and $\vec{u}'\vec{\cdot}\vec{\nabla}T'$ terms from the evolution equations for the perturbations.
Substituting \eqss{EQ:reynolds_dc_u}{EQ:reynolds_dc_p} into \eqss{EQ:motion1}{EQ:motion3} followed by subtracting \eqs{EQ:T0_IVP}{EQ:p_bar} from the resulting temperature and $\hat{z}$ momentum equations gives
\begin{align}
    \nabla \cdot \vec{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\vec{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \vec{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \overline{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad \vec{u'}|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
This is now a linear problem in $\vec{u}'$ and $T'$ which can be solved as an eigenvalue problem.

@@

For each $k_x$, we can assess the stability of the perturbations by solving for the eigenvalue $s$, whose imaginary component $\sigma$ plays the role of an exponential growth rate. 
Positive eigenvalues indicate that the system is unstable to small disturbances of wavenumber $k_x$, while negative eigenvalues indicate stability. 
A complete linear stability analysis requires solution over the full spectrum of wavenumbers. 
The prototypical case is used to demonstrate that the critical Rayleigh number $\Ra_c = 1708$ when $\frac{\partial \overline{T}}{\partial z} = -1$.

To calculate the advective heat flux in equation~\ref{EQ:T0_IVP}, we can sum $\langle w'T' \rangle_x$ from each horizontal wavenumber individually.
In this way, the heat flux from the perturbations influence the evolution of $\overline{T}$.
But the evolution of $\overline{T}$ also influences the perturbations, as equation~\ref{EQ:linear3} depends on $\partial_z \overline{T}$.
Thus, the mean temperature and perturbation fields are coupled.

\section{Perturbation Evolution}\label{sec:evolution}
The linearized system \eqss{EQ:linear1}{EQ:linear3} does not constrain the amplitude of the eigenmodes, $A$.
However, the advective heat flux is proportional to $A^2$, so we need to specify the amplitude in order to solve equation~\ref{EQ:T0_IVP}.
To evolve $\overline{T}$, we assume the perturbations evolve on a much faster time-scale than the mean temperature, as in \cite{michel_chini_2019}.
Stable modes $(\sigma < 0)$ decay away rapidly. 
Unstable modes $(\sigma > 0)$ will not persist on the slow time-scale because the advective term $\langle w'T' \rangle_x$ tends to stabilize $\overline{T}$, thereby creating a negative feedback loop.
Only marginally stable modes can be maintained on the slow time-scale.
Therefore the amplitude $A$ must satisfy
\begin{equation}
    \max_{k_x} \{ \sigma \} = 0.
\end{equation}

For various $\Ra$ and fixed $\Pr = 1$, we seek marginally-stable thermal equilibria (MSTE) satisfying $\frac{\partial \overline{T}}{\partial t} = 0$ according to \eq{EQ:T0_IVP}. 
We employ the \texttt{Dedalus} pseudo-spectral python framework\footnotemark[1] \cite{Dedalus_2020} to solve the eigenvalue problem outlined in Section \ref{sec:model} as well as the time evolution equations \eq{EQ:T0_IVP}.
We represent each field with Chebyshev polynomials and use the 3/2 dealiasing rule to calculate the advective heat flux.
The necessary number of basis functions increases with $\Ra$ as the eigenfunctions include increasingly small-scale features (see Appendix \ref{sec:tables}). 
We use the \texttt{Eigentools} package\footnotemark[2] \cite{Eigentools} to manipulate the eigenfunctions and calculate the advective heat flux $\langle w' T' \rangle_x$.

\footnotetext[1]{We employ \texttt{Dedalus} version v2.2006

\url{https://github.com/DedalusProject/dedalus/releases}}

\footnotetext[2]{We employ \texttt{Eigentools} commit

f457afc3193c32c16fdff94cdad962d07edea56b

\url{https://github.com/jsoishi/eigentools}}

To initialize our calculations, we construct a marginally-stable initial temperature profile $\overline{T}(z, t=0)$ whose equation is given in appendix~\ref{sec:initial_profile}. 
We use \eq{EQ:T0_IVP} to evolve $\overline{T}(z, t_0)$ into a new marginally-stable profile $\overline{T}(z, t_0 + \Delta t)$.
We use a second-order, two-stage IMEX Runge-Kutta method.
The eigenfunctions and amplitudes are assumed to be constant over the timestep.
We symmetrize $\overline{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each timestep.
This ensures $\overline{T}(z) = -\overline{T}(-z)$.
On every timestep, we calculate the marginally stable eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $\Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\overline{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing $\sigma$. 
    Advection tends to stabilize the system, decreasing $\sigma$. 
    We find the eigenfunction amplitude $A^2$ such that the combination of diffusion and advection yields a new, ``evolved,'' marginally-stable mean temperature profile $\overline{T}(z, t_0 + \Delta t)$. In this case $A^2 \approx 1$ due to approximately equal magnitudes of the diffused and advected eigenvalues.}
    \label{fig:iteration_spectra} 
\end{figure}

Consider a marginally-stable temperature profile $\overline{T}(z, t_0)$.
By definition, its maximum growth rate is 0.
Diffusing $\overline{T}(z, t_0)$ tends to increase its growth rates while ignoring the diffusive term and evolving according to advection tends to decrease its growth rates.
$A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new IVPs
\begin{align}\label{eqn:diffusion_only}
    &\frac{\partial \overline{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \overline{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \overline{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = 0 
\end{align}
where $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ denote to the diffused and advected temperature profiles.
We initialize $\overline{T}_{\rm{diff}}$ and $\overline{T}_{\rm{adv}}$ with $\overline{T}(z, t_0)$, and evolve the equations to $t=t_0+\Delta t$.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then, using the diffused and advected profiles, we compute the eigenvalues $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$ and $t=t_0+\Delta t$.
A preliminary approximation $\Lambda^2_0$ of the appropriate amplitude $A^2$ is given by
\begin{equation}
    A^2 \approx \Lambda^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2$ and $\Lambda^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.

Using the approximate amplitude $\Lambda_0^2$, the background temperature $\overline{T}(z, t_0)$ is then evolved to $t=t_0+\Delta t$ according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
This background temperature is typically close to, but not exactly, marginally stable.
We use Newton's method to find an amplitude $A^2$ such that the maximum growth rate is zero to within $10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e. $\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In section \ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.

@@

Our estimate $\vec{\Lambda^2_0}$ does not give a background temperature for which all $N$ modes are exactly marginally stable.
We refine our vector of amplitudes via Newton's method.
This requires calculating the Jacobian matrix 
\begin{equation}
    J = \begin{bmatrix}
        \nabla \sigma_1 (A_1, A_2, ..., A_N) \\
        \nabla \sigma_2 (A_1, A_2, ..., A_N) \\
        \vdots \\
        \nabla \sigma_N (A_1, A_2, ..., A_N) 
    \end{bmatrix}.
\end{equation}
We approximate $J$ via first-order finite differences, which requires an additional $N^2$ sparse eigenvalue solves.
%A once-refined estimate $\vec{A^2_1}$ is then given by
%% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
%\begin{equation}
%    \vec{\tilde{A}^2} \approx \vec{A^2_1} = -J^{-1} \vec{\sigma}(\vec{A^2_0}).
%\end{equation}
%We then use each subsequent guess to adjust the Jacobian via Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We iterate Newton's method until all marginally stable modes have eigenvalues within $10^{-9}$ of zero.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the coupling of $\langle w'T' \rangle_x$ with $\overline{T}$.
Over the course of a large time step, $\overline{T}$ evolves according to \eq{EQ:T0_IVP} and eventually the original eigenfunctions cease to provide a stabilizing influence.
This limits the timestep size we can take.
